---
title: "R Notebook"
output:
  md_document:
    variant: markdown_github
  pdf_document: default
  html_document:
    df_print: paged
---

# Basic Commands

## Hello World

```{r}
print("Hello World")
```

## Get help for a command

```{r}
?load
```

## Calculations

R performs addition, subtraction, multiplication, and division with +, -, \*, and /

```{r}
25 * 4 + 9 / 3
```

## Comments

R interprets anything after a \# as a comment.

```{r}
persnickety_count <- 0
```

## Data Types

-   Numeric: Any number with or without a decimal point: 23, 0.03and the numeric null value NA.
-   Character: Any grouping of characters on your keyboard (letters, numbers, spaces, symbols, etc.) or text. Most strings are surrounded by single quotes: ' ... ' or double quotes " ... ", though we prefer single quotes. Sometimes you will hear this type referred to as "string."
-   Logical: This data type only has two possible values--- either TRUE or FALSE (without quotes). It's helpful to think of logical types or booleans as on and off switches or as the answers to a "yes" or "no" question.
-   Vectors: A list of related data that is all the same type.
-   NA: This data type represents the absence of a value, and is represented by the keyword NA (without quotes) but it has its own significance in the context of the different types. That is there is a numeric NA, a character NA, and a logical NA.

```{r}
class(2)        # numeric
class('hello')  # character
class('23')     # character
class (FALSE)   # logical
class(NA)       # logical
```

## Variables

```{r}
# Greeting
message_string <- "Hello there"
print(message_string)
 
# Farewell
message_string <- "Hasta la vista"
print(message_string)

result <- 1 + 2
print(result)
```

## Vectors

In R, vectors are a list-like structure that contain items of the same data type.

```{r}
phone <- c(123, 456, 789)
print(phone)
typeof(phone)
print(phone[1])   # R starts to count at 1

```

## Conditionals

```{r}
if (TRUE) {
   print("Go to sleep!")
} else {
   print("Wake up!")
}
```

### ifelse() function

This function takes a logical expression test on some vector, if the test is True, the first action True is taken otherwise the second action False is taken. 

```{r}
# ifelse() function
score <- c(77,35,89,100,45,67,50)
ifelse(score >= 50, "Pass", "Fail")

```

## Comparison Operators

-   Less than: `<`

-   Greater than: `>`

-   Less than or equal to: `<=`

-   Greater than or equal to: `>=`

-   Is equal to: `==`

-   Is NOT equal to: `!=`

```{r}
10 < 12 # Evaluates to TRUE
```

## Logical Operators

-   the AND operator (`&`)

-   the OR operator (`|`)

-   the NOT operator, otherwise known as the bang operator (`!`)

```{r}
stopLight <- 'green'
pedestrians <- 0
if (stopLight == 'green' & pedestrians == 0) {
  print('Go!');
} else {
  print('Stop');
}
```

## Calling a Function

```{r}
sort(c(2,4,10,5,1)); # Outputs c(1,2,4,5,10)
length(c(2,4,10,5,1)); # Outputs 5
sum(5,15,10) #Outputs 30

data <- c(120,22,22,31,15,120)

unique_vals <- unique(data)
print(unique_vals)

solution <- sqrt(49)
print(solution)

round_down <- floor(3.14)
print(round_down)

round_up <- ceiling(3.14)
print(round_up)
```

## Loops

### For-Loop

```{r}
for ( x in 1:5){
    print(x)
}
```

### While-Loop

```{r}
x  <-  10
while ( x >= 5 ) {
     print (x)
     x  <-  x-1
}
```

### Repeat-Loop

```{r}
counter <- 10

repeat { 
    print(counter)
    counter <- counter - 1
    if (counter < 5)
        break
}
```

## Recursion

```{r}
factorial <- function(n) {
    if (n == 0)     
        return (1)

    else 
        return (n * factorial (n-1))
}

factorial(5)
```

## Importing Packages

Base R refers the R language by itself and all that it can do without importing any packages.

[CRAN](https://cran.r-project.org/) offers more packages to extend the functionality of R. Dplyr is a package used to clean, process, and organize data which you will use as you learn about R. The goal of 'readr' is to provide a fast and friendly way to read rectangular data (like 'csv', 'tsv', and 'fwf').

```{r}
# load libraries
library('dplyr')
library('readr')
```

## CSV Import options

```{r}
data <- read.csv('data/biostats.csv')
# data <- read.csv(file.choose())
data <- read.csv("https://people.sc.fsu.edu/~jburkardt/data/csv/biostats.csv")
```

### readr

```{r}
# load data frame
biostats <- read_csv('data/biostats.csv')

# inspect data frame
biostats
head(biostats)
summary(biostats)

```

## Averages (Lageparameter)

### Mean (Mittelwert)

It is calculated by taking the sum of the values and dividing with the number of values in a data series. The function **mean()** is used to calculate this in R.

    mean(x, trim = 0, na.rm = FALSE, ...)

-   **x** is the input vector.

-   **trim** is used to drop some observations from both end of the sorted vector. e.g. trim = 0.3 drops 3 values from start and end.

-   **na.rm** is used to remove the missing values from the input vector. If there are missing values, then the mean function by default returns NA.

```{r}
# Create a vector. 
x <- c(12,7,3,4.2,18,2,54,-21,8,-5)

# Find Mean.
result.mean <- mean(x)
print(result.mean)
```

### Median

The middle most value in a data series is called the median. The **median()** function is used in R to calculate this value.

    median(x, na.rm = FALSE)

-   **x** is the input vector.

-   **na.rm** is used to remove the missing values from the input vector.

```{r}
# Create the vector.
x <- c(12,7,3,4.2,18,2,54,-21,8,-5)

# Find the median.
median.result <- median(x)
print(median.result)
```

### Modus (Modus)

The mode is the value that has highest number of occurrences in a set
of data. Unike mean and median, mode can have both numeric and
character data.

```{r}
# Create the function.
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Create the vector with numbers.
v <- c(2,1,2,3,1,2,3,4,1,5,5,3,2,3)

# Calculate the mode using the user function.
result <- getmode(v)
print(result)

```

### Averages from a csv file

```{r}
biostats <- read_csv('data/biostats.csv')

result.mean <- mean(biostats$Age)
result.median <- median(biostats$Age)
result.mode <- getmode(biostats$Age)
print(result.mean)
print(result.median)
print(result.mode)
```

## **Statistical dispersion (Streuparameter)**

### Range (Spannweite)

The range of a set of data is the difference between the largest and smallest values. It can give you a rough idea of how the outcome of the data set will be before you look at it actually.

```{r}
x <- c(12,7,3,4.2,18,2,54,-21,8,-5)
result.range <- max(x) - min(x)
print(result.range)

```

### Interquartile range (Quartilsabstand)

Interquartile range (IQR), also called the midspread, middle 50%, or H‑spread, is a measure of statistical dispersion, being equal to the difference between 75th and 25th percentiles, or between upper and lower quartiles.

-   **x** is a numeric vector.

-   **na.rm** Should missing values be removed?

-   **type** is an integer between 1 and 9 selecting one of the nine quantile algorithms.

```{r}
x <- c(12,7,3,4.2,18,2,54,-21,8,-5)
result.iqr <- IQR(x, na.rm = FALSE, type = 7)
print(result.iqr)
```

### Variance (Varianz)

Variance is the expectation of the squared deviation of a random variable from its mean.

```{r}
v <- c(2,1,2,3,1,2,3,4,1,5,5,3,2,3)
result.var <- var(v)
print(result.var)
```

### Standard deviation (Standardabweichung)

The standard deviation is a measure of the amount of variation or dispersion of a set of values.

```{r}
v <- c(2,1,2,3,1,2,3,4,1,5,5,3,2,3)
result.sd <- sd(v)
print(result.sd)
```

### **Statistical dispersion from csv file**

```{r}
biostats <- read_csv('data/biostats.csv')

result.range <- max(biostats$Age) - min(biostats$Age)
result.iqr <- IQR(biostats$Age, na.rm = FALSE, type = 7)
result.var <- var(biostats$Age)
result.sd <- sd(biostats$Age)
print(result.range)
print(result.iqr)
print(result.var)
print(result.sd)
```
